---
title: Memory phenotypes of CD4+ T cells after acute infection
date: "`r Sys.Date()`"
author: "M. Andreatta, T. Ciucci, S. Carmona"
output:
  rmdformats::readthedown:
    self-contained: true
    highlight: haddock
    thumbnails: false
    css: styles.css
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file, encoding=encoding,
  output_file=file.path(dirname(input_file), out_dir, 'Kunzli.html'))})
---

```{r setup, echo=FALSE}
#install.packages("rmdformats")
#Template markdown setup
library(knitr)
library(rmdformats)
library(formatR)

## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               cache.lazy=FALSE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE,
               dev='png')
opts_knit$set(width=75)

```

# Background

CD4+ T cells play a central role in initiating and shaping adaptive immune responses. After pathogen control or clearance, the vast majority of activated CD4 T cells undergo apoptosis, but a subset remain long term, forming a population of long-lived memory cells. These memory cells will be able to respond more effectively following a secondary infection. 

Differentiation into a specific CD4+ T cell subset depends on several factor, e.g. the tissue, the antigen, and which cytokines are found in the surrounding environment. One such subset, follicular helper T cells (Tfh), are located in secondary lymphoid organs and promote antibody responses via direct cell contact with B cells. However, long-lived Tfh cells have been difficult to detect and characterize. In their study, [Kunzli et al. (2020)](https://pubmed.ncbi.nlm.nih.gov/32144185/) found that Tfh cells are particularly susceptible to NAD-induced cell death (NICD) during sample isolation, and proposed administration of an inhibitor blocking NICD to enrich Tfh cell populations after infection. We will re-analyze their single-cell data for NICD-treated cells at day 35 (versus a control sample at the same time point) using an automated pipeline.

For this purpose, we will apply ProjecTILs to analyse these samples in the context of a reference atlas of CD4+ T cells in viral infection (described in [Andreatta et al. (2022)](https://doi.org/10.7554/eLife.76339)). This atlas can also be explored interactively through the SPICA portal at: [https://spica.unil.ch/refs/viral-CD4-T](https://spica.unil.ch/refs/viral-CD4-T)


# R Environment

```{r, message=FALSE, warning=F, results=FALSE}
library(renv)
renv::restore()

library(ggplot2)
library(reshape2)
library(patchwork)
library(ProjecTILs)
```

# scRNA-seq data preparation

We can download the single-cell data directly from Gene Expression Omnibus (GEO).
```{r, message=FALSE, warning=F, results=FALSE}
library(GEOquery)
geo_acc <- "GSE134157"
datadir <- "input/Kunzli"

gse <- getGEO(geo_acc)

system(paste0("mkdir -p ", datadir))
getGEOSuppFiles(geo_acc, baseDir = datadir)

exp_mat.1 <- read.table(sprintf("%s/GSE134157/GSE134157_UMImatrix_NICD_protector.tsv.gz", datadir))
exp_mat.2 <- read.table(sprintf("%s/GSE134157/GSE134157_UMImatrix_no_NICD_protector.tsv.gz", datadir))
```

Genes in these matrices are given by Ensembl ID - we will convert them to gene names.
```{r}
dataUrl <- "https://drive.switch.ch/index.php/s/iJKbWGHwOhY1Llu/download"
fname <- sprintf("%s/mart_conversion_Mm.txt", datadir) 

download.file(dataUrl, fname)

table <- read.csv(fname,sep='\t')

ID2name <- table$Gene.name
names(ID2name) <- table$Gene.stable.ID
ID2name <- ID2name[!duplicated(names(ID2name))]

#Convert rownames in gene matrices
exp_mat.1 <- exp_mat.1[rownames(exp_mat.1) %in% names(ID2name),]
rownames(exp_mat.1) <- ID2name[rownames(exp_mat.1)]

exp_mat.2 <- exp_mat.2[rownames(exp_mat.2) %in% names(ID2name),]
rownames(exp_mat.2) <- ID2name[rownames(exp_mat.2)]

```

Create Seurat objects from the sigle-cell data. Here we also choose to downsample to 1000 cells per sample, to balance the contribution of each sample to the analysis.
```{r}
query.list <- list()

query.list[["protector"]] <- CreateSeuratObject(counts = exp_mat.1, min.cells = 3, min.features = 50)
query.list[["protector"]]$Sample <- substring(colnames(query.list[["protector"]]),18)
query.list[["protector"]]$condition <- "protector"

query.list[["control"]] <- CreateSeuratObject(counts = exp_mat.2, min.cells = 3, min.features = 50)
query.list[["control"]]$Sample <- substring(colnames(query.list[["control"]]),18)
query.list[["control"]]$condition <- "control"

query.merged <- merge(query.list[[1]], query.list[[2]])

#Downsample to 1000 cells per sample
set.seed(1234)
Idents(query.merged) <- "Sample"
query.merged <- subset(query.merged, cells=WhichCells(query.merged,downsample = 1000))
table(query.merged$Sample)

```

# Load reference atlas

The CD4+ T cell atlas is described in [Andreatta et al. (2022)](https://doi.org/10.7554/eLife.76339), and can be downloaded from Figshare at: [doi.org/10.6084/m9.figshare.16592693.v2](https://doi.org/10.6084/m9.figshare.16592693.v2)

You can also use the commands below to download the atlas directly within R, and load it into memory.

```{r fig.height=3.5, fig.width=5}
#Download the reference atlas
cd4.atlas.file <- "ref_LCMV_CD4_mouse_release_v1.rds"
if (!file.exists(cd4.atlas.file)) { 
  dataUrl <- "https://figshare.com/ndownloader/files/31057081"
  download.file(dataUrl, cd4.atlas.file)
}
ref <- load.reference.map(cd4.atlas.file)
DimPlot(ref, label = T, cols = ref@misc$atlas.palette)
```

# Dataset projection

Next, we apply ProjecTILs to map the new datasets to the reference atlas. If enough cells (>500-1000) are present in individual samples, we recommend projecting them individually into the reference atlas, to fully benefit from ProjecTILs' batch-effect correction.

```{r}
query.by.sample <- SplitObject(query.merged, split.by = "Sample")

query.projected <- make.projection(query.by.sample, ref=ref)  
```


Visualize the projection of individual samples. Note the distribution of cell subtypes in the barplots below the UMAPs.
```{r fig.width=12, fig.height=6}
plots <- list()
palette <- ref@misc$atlas.palette

for (i in seq_along(query.projected)) {
  sample <- names(query.projected)[i]
  cond <- unique(query.projected[[i]]$condition)
  
  query.projected[[i]] <- cellstate.predict(ref=ref, query=query.projected[[i]], reduction = "umap", ndim=2)
  
  
  plots[[i]] <- plot.projection(ref, query.projected[[i]], linesize=0.5, pointsize=0.5) + 
    ggtitle(paste(sample,cond)) + NoLegend()  + 
    theme(legend.position = "none", panel.grid = element_blank(),
        axis.title = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank())
  
    plots[[i+3]] <- plot.statepred.composition(ref, query=query.projected[[i]], cols=palette, metric="Percent") + 
      ggtitle(" ") + ylim(0,60)  + theme_bw() + 
      theme(panel.grid = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x =  element_blank())
}

g <- wrap_plots(plots, ncol=3)
plot(g)

```

Radar plots can be useful as a quality control, to ensure that expression profile for the projected cells match in a reasonable manner the profiles of reference subtypes.
```{r fig.height=9, fig.width=13}
features <-c("Ifng","Ccl5","Gzmb","Cxcr6","Selplg","Id2","Tbx21",
             "Ly6c2","Cxcr5","Tox","Tox2","Izumo1r","Pdcd1","Tnfsf8",
             "Ccr7","Il7r","Tcf7","Eomes","Ifit1")

plot.states.radar(ref, query=query.projected, 
                  min.cells=30, genes4radar = features)

```

To summarize the effect of the NICD-protector vs. Control, we can pool samples from the same condition and view them jointly.
```{r}
query.bycondition <- list()

query.bycondition[['Control']] <- query.projected$Spleen_0
query.bycondition[['Protector']] <- ProjecTILs:::merge.Seurat.embeddings(query.projected$Spleen_1, query.projected$Spleen_2)
```

Projection plots and composition plots by condition, after pooling.
```{r fig.width=9, fig.height=6}
plots <- list()

for (i in seq_along(query.bycondition)) {
  cond <- names(query.bycondition)[i]

  plots[[i]] <- plot.projection(ref, query.bycondition[[i]], linesize=0.5, pointsize=0.5) + 
    ggtitle(cond) + NoLegend() 
  
  query.bycondition[[i]] <- cellstate.predict(ref=ref, query=query.bycondition[[i]], reduction = "umap", ndim=2)
  
  plots[[i+2]] <- plot.statepred.composition(ref, query=query.bycondition[[i]], cols=palette, metric="Percent") + 
    ggtitle(" ") + ylim(0,60)  + theme_bw() + 
        theme(panel.grid = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x =  element_blank())
}

g <- wrap_plots(plots, ncol=2)
g
```

In line with the original publication, projection of CD4+ T cell scRNA-seq data into the reference atlas show that Tfh-memory cells are enriched after NICD-protector treatment, compared to control. Importantly, ProjecTILs enables analyzing complex single-cell data in minutes without specific expertise in T cell biology.

# References

* M. Künzli, D. Schreiner, T. C. Pereboom, N. Swarnalekha, ..., C. G. King **"Long-lived T follicular helper cells retain plasticity and help sustain humoral immunity"** (2020) Science Immunology
* M. Andreatta, Z. Sherman, A. Tjitropranoto, M. C. Kelly, T. Ciucci, S. J. Carmona **"A single-cell reference map delineates CD4+ T cell subtype-specific adaptation during acute and chronic viral infections"** eLife 2022 doi: https://doi.org/10.7554/eLife.76339


